---
documentclass: jss
author:
  - name: Fabien Girka
    orcid: 0000-0003-2843-1104
    affiliation: |
      | Laboratoire des Signaux et Systèmes
    address: >
      Université Paris-Saclay, CNRS, CentraleSupélec, \newline 
      Gif-sur-Yvette, 91190, France \newline
      Institut du Cerveau de Paris (ICM), Sorbonne Université \newline 
      Assistance Publique-Hôpitaux de Paris (AP-HP)
    email: \email{fabien.girka@centralesupelec.fr}
  - name: Etienne Camenen
    affiliation: | 
      | INSERM
    address: >
      Inserm U1127, CNRS UMR 7225, \newline
      Hôpital Universitaire Pitié-Salpêtrière \newline 
      75013 Paris, France \newline
      Institut du Cerveau de Paris (ICM), Sorbonne Université \newline 
      Assistance Publique-Hôpitaux de Paris (AP-HP)
  - name: Caroline Peltier
    affiliation: 'INRAE \AND'
    affiliation2: 'INRAE'
    address: >
      Centre des Sciences du Goût et de l’Alimentation \newline
      CNRS, INRAE, Institut Agro \newline
      University of Bourgogne F-21000 Dijon \newline
      CNRS, INRAE, PROBE research infrastructure \newline 
      ChemoSens facility, F-21000 Dijon, France \newline
      Institut du Cerveau de Paris (ICM), Sorbonne Université \newline 
      Assistance Publique-Hôpitaux de Paris (AP-HP)
  - name: Arnaud Gloaguen
    affiliation: |
      | CEA 
    address: >
      Centre National de Recherche en Génomique Humaine \newline 
      Institut de Biologie François Jacob \newline 
      CEA, Université Paris-Saclay, Évry, France 
  - name: Vincent Guillemot
    affiliation: 'Institut Pasteur \AND'
    affiliation2: 'Institut Pasteur'
    address: >
      Université Paris Cité \newline
      Bioinformatics and Biostatistics Hub \newline 
      F-75015 Paris, France
  - name: Laurent Le Brusquet
    affiliation: |
      | Laboratoire des Signaux et Systèmes
    address: >
      Université Paris-Saclay, CNRS, CentraleSupélec \newline 
      Gif-sur-Yvette, 91190, France 
  - name: Arthur Tenenhaus
    orcid: 0000-0003-3459-1518
    affiliation: |
      | Laboratoire des Signaux et Systèmes
    address: >
      Université Paris-Saclay, CNRS, CentraleSupélec \newline 
      Gif-sur-Yvette, 91190, France \newline
      Institut du Cerveau de Paris (ICM), Sorbonne Université \newline 
      Assistance Publique-Hôpitaux de Paris (AP-HP)
    email: \email{arthur.tenenhaus@centralesupelec.fr}
    # use a different affiliation in adress field (differently formated here)
title:
  formatted: "Multiblock data analysis with the RGCCA package"
  plain: "Multiblock data analysis with the RGCCA package"
  short: "Regularized Generalized Canonical Correlation Analysis"
pagetitle: "Multiblock data analysis with the RGCCA package"
abstract: 
    Multiblock component methods aim to study the relationships between several sets of variables. Regularized Generalized Canonical Correlation Analysis (RGCCA) is a unified and flexible framework that gathers six decades of multiblock component methods. RGCCA offers a unified implementation strategy for all these methods. This implementation is made available within the RGCCA package. In addition, the RGCCA package produces graphical outputs and statistics to assess the robustness/significance of the analysis. The usefulness of the RGCCA package is illustrated in this paper on two real datasets. The RGCCA package is freely available on the ComprehensiveR Archive Network (CRAN) http://www.r-project.org/ and GitHub https://github.com/rgcca-factory/RGCCA.
keywords:
  formatted: [Multiblock component methods, RGCCA, data integration]
  plain:     [Multiblock component methods, RGCCA, data integration]
preamble: >
  \usepackage{amsmath}
  \usepackage{amsfonts}
  \usepackage{times}
  \usepackage{bm}
  \usepackage{soul}
  \usepackage{epsfig}
  \usepackage{amssymb}
  \usepackage{lscape}
  \usepackage{float}
  \usepackage{latexsym}
  \usepackage{graphicx,psfrag,color}
  \usepackage{multirow} 
  \usepackage[space]{grffile}
  \usepackage{amsthm}
  \usepackage{enumerate}
  \usepackage{enumitem}
  \usepackage{setspace}
  \usepackage{subfigure}
  \usepackage{longtable}
  \usepackage{etoolbox} 
  \usepackage{pdfpages}
  \usepackage[mathscr]{euscript}
  \usepackage[T1]{fontenc}
  \usepackage[misc]{ifsym}
  \usepackage{wasysym}
  \usepackage{hyperref}
  \usepackage[width=\textwidth]{caption}
  \usepackage{algorithmic, algorithm}
output: rticles::jss_article
bibliography: biblio.bib
vignette: >
  %\VignetteIndexEntry{RGCCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
  extension: pdf
---


```{=tex}
\newtheorem{theorem}{theorem}[section]%
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{remark}[theorem]{Remark}
```



```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
options(ggrepel.max.overlaps = Inf)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})

knitr::opts_chunk$set(
  fig.path = "figures/"
)
```

# Introduction

A challenging problem in multivariate statistics is to study relationships 
between several sets of variables measured on the same set of individuals. 
This paradigm is referred to by several names, including "learning from 
multimodal data", "data integration", "multiview data", "multisource data", 
"data fusion" or "multiblock data analysis". Despite the availability of various 
statistical methods and dedicated software for multiblock data analysis, handling 
multiple highly multivariate datasets remains a critical issue for effective 
analysis and knowledge discovery.

Regularized Generalized Canonical Correlation Analysis (RGCCA) \cite{Tenenhaus2011, Tenenhaus2014, Tenenhaus2015, Tenenhaus2017} is a unified 
statistical framework that gathers six decades of multiblock component methods.
While RGCCA encompasses a large number of methods, it is based on a single optimization problem. This unified approach bears immediate practical implications, facilitating statistical analyses and implementation strategies.

In this paper, we present the R package called $\small{\texttt{RGCCA}}$ (Girka et al, 2023), which implements the RGCCA framework. In addition to the main statistical functionalities, the RGCCA package offers various advanced ggplot visualization tools to help users interpret and extract meaningful information from integrated data. It also includes metrics for assessing the robustness and significance of the analysis.

For the statistical computing environment R \cite{R2022}, various R packages are 
have come to the fore for carrying out multiblock data analysis. The choice of 
the package is dictated mostly by the domain of application.
The bioinformatician community mostly uses $\small{\texttt{mixOmics}}$ \cite{Rohart2017} which wraps the main functions of the RGCCA package for performing multiblock analyses.

The ecological community widely uses $\small{\texttt{ade4}}$ 
(\cite{Dray2007}), which mainly relies on three approaches 
for performing multiblock analysis: Multiple co-inertia analysis (MCOA) \cite{Chessel1996} or Multiple Factor Analysis (MFA) \cite{Escofier1994} and Statis \cite{Lavit1994}. 

$\small{\texttt{FactoMineR}}$ \cite{Le2008} is also one widely used package for multiblock methods mainly due to the implementation of MFA and, to a somewhat lesser extent, Generalized Procrustes Analysis (GPA) \cite{Gower1975}.

The $\small{\texttt{multiblock}}$ package \cite{Liland2022} covers a wide range of multiblock methods for data fusion, but intensively relies on a wrapper strategy for performing multiblock analyses. Consequently, this results in strong dependencies  between the $\small{\texttt{multiblock}}$ package and other packages including (i) FactoMineR (for MFA, GPA), RGCCA (for Hierarchical PCA \citep{Wold1996, Hanafi2010}, MCOA, Interbattery Factor Analysis \citep{Tucker1958} , Carroll's GCCA \cite{Carroll1968a}), ade4 (Multiblock redundancy analysis \cite{Bougeard2011}, statis), r.jive (for JIVE \cite{Lock2013}), RegularizedSCA (for DISCO \cite{Schouteden2014}).

To the sake of completeness, we also mention the $\small{\texttt{multiview}}$ R package that can be used for predicting a univariate response (member of the exponential family of distributions) from several blocks of variables \cite{}Ding2022}.

Each package uses its own way of specifying multiblock models and storing the results.

For Python users, $\small{\texttt{mvlearn}}$ \cite{Perry2021} seems to be the most advanced Python module for multiview data. mvlearn offers a suite of algorithms for learning latent space embeddings and joint representations of views including a limited version of the RGCCA framework, a kernel versions of GCCA (\cite{Hardoon2004, \cite{Bach2002}, Deep CCA \cite{Andrew2013}. Several other methods for dimensionality reduction and joint subspace learning are included as well, such as multiview 
multidimensional scaling \cite{Trendafilov2010}.

From this perspective, it appears that the $\small{\texttt{RGCCA}}$ package already plays a central role within the R community for conducting multiblock analyses. 
Also, special attention has been paid to ensuring that $\small{\texttt{RGCCA}}$'s implementation of various multiblock component methods, including MCOA and MFA, aligns with results obtained from other packages such as $\small{\texttt{ade4}}$ or $\small{\texttt{FactoMineR}}$.

Within the $\small{\texttt{RGCCA}}$ package, all implemented methods share the 
same function interface and a clear class structure. In addition, the 
$\small{\texttt{RGCCA}}$ package relies on the single $\small{\texttt{rgcca()}}$ 
function for specifying the multiblock models. It also provides  several 
utility functions for data preprocessing and several plots 
for diagnostics or visualization of the results from multiblock analysis.  
In addition, the package includes several built-in datasets and examples to 
help users get started quickly. Package $\small{\texttt{RGCCA}}$ is available 
from the Comprehensive R Archive Network (CRAN), at https://CRAN.R-project.org/package=RGCCA and can be installed from the R console with the following command:

```{r, eval = FALSE}
install.packages("RGCCA")
```

In this paper, we present an overview of the RGCCA framework's theoretical 
foundations, summarize the optimization problem under which all the algorithms 
were designed, and provide code examples to illustrate the package's usefulness 
and versatility. We believe that our package provides a valuable contribution 
to the field of multiblock data analysis and will enable researchers to conduct 
more effective analyses and gain new insights into complex datasets.

The paper's remaining sections are organized as follows: Section 2 summarizes the theoretical foundations of the RGCCA framework. Section 3 presents the general optimization problem and the master algorithm underlying the RGCCA framework. Sections 4 and 5 provide code examples to illustrate the package's capabilities. Finally, we conclude the paper in Section 6.

# Optimization background

This section presents the optimization framework under which 
all the algorithms proposed in the RGCCA framework were designed. The RGCCA 
framework gathers several methods already presented in 
\cite{Tenenhaus2011, Tenenhaus2014, Tenenhaus2015, Tenenhaus2017}. It is recalled 
here for a broader class of constraints.

The RGCCA framework relies on a master algorithm for maximizing a continuously 
differentiable multi-convex function 
$f( \mathbf a_1, \ldots, \mathbf a_J):\mathbb{R}^{p_1}\times \ldots \times \mathbb{R}^{p_J} \xrightarrow{}\mathbb{R}$ (i.e. for each $j$, $f$ is a convex function of 
$\mathbf a_j$ while all the other $\mathbf a_k$ are fixed) under the constraint that 
each $\mathbf a_j$ belongs to a compact set $\Omega_j\subset \mathbb{R}^{p_j}$. 
This general optimization problem can be formulated as follows:

\begin{align}
\underset{ \mathbf a_1, \ldots, \mathbf a_J}{\text{max}} f( \mathbf a_1, \ldots, \mathbf a_J)
\text{ s.t. }  \mathbf a_j \in \Omega_j, \text{ } j = 1, \ldots, J.
\label{master_optim}
\end{align}

For such a function defined over a set of parameter vectors $( \mathbf a_1, \ldots, \mathbf a_J)$, 
we make no difference between the notations $f( \mathbf a_1, \ldots, \mathbf a_J)$ and 
$f( \mathbf a)$, where $\mathbf a$ is the column vector 
$\mathbf a = \left(  \mathbf a_1^\top, \ldots,  \mathbf a_J^\top\right)^\top$ 
of size $p = \sum_{j=1}^{J}p_j$. Moreover, the vertical concatenation of a 
column vector is denoted $\mathbf a = \left(  \mathbf a_1; \ldots;  \mathbf a_J \right)$ 
for the sake of simplification of notation. 

## Algorithm

A simple, monotonically, and globally convergent algorithm is presented for 
solving optimization problem (\ref{master_optim}). The maximization 
of the function $f$ defined over different parameter vectors 
($\mathbf a_1, \ldots, \mathbf a_J$) is approached by updating each of the parameter 
vectors in turn, keeping the others fixed. This update rule was recommended 
in \cite{DeLeeuw1994} and is called Block Relaxation or cyclic Block Coordinate 
Ascent (BCA). 
    
Let $\nabla_j f( \mathbf a)$ be the partial gradient of $f( \mathbf a)$ 
with respect to $\mathbf a_j$. We assume $\nabla_j f( \mathbf a) \neq \mathbf{0}$ in 
this manuscript. This assumption is not too binding as 
$\nabla_j f( \mathbf a) = \mathbf{0}$ characterizes the global minimum of 
$f( \mathbf a_1 , \ldots ,  \mathbf a_J )$ with respect to $\mathbf a_j$ when the other 
vectors $\mathbf a_1 , \ldots ,  \mathbf a_{j-1} ,  \mathbf a_{j+1} , \ldots ,  \mathbf a_J$ 
are fixed. 

We want to find an update $\hat{ \mathbf a}_j\in \Omega_j$ such that 
$f( \mathbf a)\leq f( \mathbf a_1, ...,  \mathbf a_{j-1}, \hat{ \mathbf a}_j,  \mathbf a_{j+1}, ...,  \mathbf a_J)$. 
As $f$ is a continuously differentiable multi-convex function and considering 
that a convex function lies above its linear approximation at $\mathbf a_j$ for any 
$\tilde{ \mathbf a}_j\in\Omega_j$, the following inequality holds: 

\begin{equation}
\begin{gathered}
f( \mathbf a_1, ...,  \mathbf a_{j-1}, \tilde{ \mathbf a}_j,  \mathbf a_{j+1}, \ldots,  \mathbf a_J) \geq f( \mathbf a) + \nabla_jf( \mathbf a)^\top(\tilde{ \mathbf a}_j -  \mathbf a_j).
\label{minorizing_ineq}
\end{gathered}
\end{equation}

On the right-hand side of the inequality (\ref{minorizing_ineq}), 
only the term $\nabla_jf( \mathbf a)^\top\tilde{ \mathbf a}_j$ is relevant to 
$\tilde{ \mathbf a}_j$ and the solution that maximizes the minorizing function 
over $\tilde{ \mathbf a}_j\in\Omega_j$ is obtained by considering the following 
optimization problem:

\begin{equation}
\hat{ \mathbf a}_j = \underset{\tilde{ \mathbf a}_j\in\Omega_j}{\text{argmax }} \nabla_j f( \mathbf a)^\top \tilde{ \mathbf a}_j := r_j( \mathbf a).
\label{core_update}
\end{equation}

The entire algorithm is subsumed in Algorithm \ref{master_algo}.

\begin{algorithm}[!ht]
	\caption{Algorithm for the maximization of a continuously differentiable multi-convex function}
	\begin{algorithmic}[1]
		\STATE {\bfseries Result:} {$\mathbf a_1^s, \ldots,  \mathbf a_J^s$ (approximate solution of (\ref{master_optim}))}
		\STATE {\bfseries Initialization:} {choose random vector $\mathbf a_j^0\in\Omega_j, j =1, \ldots, J$, $\varepsilon$;}
		\STATE$s = 0$ ;
		\REPEAT
		\FOR{$j=1$ {\bfseries to} $J$}
		\STATE \hspace{-2cm}$\vcenter{\begin{equation}
		     \mathbf a_j^{s+1} = r_j\left(  \mathbf a_1^{s+1}, \ldots,  \mathbf a_{j-1}^{s+1},  \mathbf a_j^{s}, \ldots,  \mathbf a_J^{s}\right).
		\end{equation}}$
		\ENDFOR
		\STATE$s = s + 1$ ;
		\UNTIL{$f( \mathbf a_1^{s+1}, \ldots,  \mathbf a_J^{s+1})-f( \mathbf a_1^s, \ldots,  \mathbf a_J^s) < \varepsilon$}
	\end{algorithmic}
	\label{master_algo}
\end{algorithm}

We need to introduce some extra notations to present the convergence 
properties of Algorithm  \ref{master_optim}: 
$\Omega = \Omega_1 \times \ldots \times \Omega_J$, 
$\mathbf a = \left( \mathbf a_1; \ldots; \mathbf a_J\right) \in \Omega$, 
$c_j \text{ : } \Omega\mapsto\Omega$ is an operator defined as 
$c_j( \mathbf a) = \left( \mathbf a_1; \ldots;  \mathbf a_{j-1} ; r_j( \mathbf a) ;  \mathbf a_{j+1} ; \ldots;  \mathbf a_J\right)$ 
with $r_j( \mathbf a)$ introduced in equation (\ref{core_update}) and 
$c \text{ : } \Omega\mapsto\Omega$ is defined as 
$c = c_J\circ c_{J-1}\circ ... \circ c_1$, where $\circ$ stands for the 
composition operator. Using the operator $c$, 
the \guillemotleft for loop\guillemotright{} inside Algorithm 
\ref{master_optim} can be replaced by the following recurrence 
relation: $\mathbf a^{s+1} = c( \mathbf a^s)$. The convergence properties of Algorithm 
\ref{master_optim} are summarized in the following proposition:

\begin{proposition}
	Let $\left\lbrace  \mathbf a^s\right\rbrace_{s=0}^{\infty}$ be any sequence 
	generated by the recurrence relation $\mathbf a^{s+1} = c( \mathbf a^s)$ with 
	$\mathbf a^0\in\Omega$. Then, the following properties hold:
	\begin{enumerate}[topsep=0pt,itemsep=-0.75ex,partopsep=1ex,parsep=1ex, label = {(\alph*)}]
		\item  \label{prop_pt1} The sequence $\left\lbrace f( \mathbf a^s)\right\rbrace $ is monotonically increasing and therefore convergent as $f$ is bounded on $\Omega$. This result implies the monotonic convergence of Algorithm \ref{master_algo}.
		\item  \label{prop_pt2} If the infinite sequence $\left\lbrace f( \mathbf a^s)\right\rbrace $ involves a finite number of distinct terms, then the last distinct point satisfies $c( \mathbf a^s) =  \mathbf a^s$ and therefore is a stationary point of problem \ref{master_algo}. 
    \item  \label{prop_pt3} $\underset{s\xrightarrow[]{}\infty}\lim{f( \mathbf a^s) = f( \mathbf a)}$, where $\mathbf a$ is a fixed point of $c$.
		\item  \label{prop_pt4} The limit of any convergent subsequence of $\left\lbrace  \mathbf a^s\right\rbrace $ is a fixed point of $c$.
		\item  \label{prop_pt5} The sequence $\left\lbrace  \mathbf a^s \right\rbrace $ is asymptotically regular: $\underset{s\xrightarrow[]{}\infty}\lim{\sum_{j=1}^{J} \Vert  \mathbf a_j^{s+1} -  \mathbf a_j^s \Vert} = 0$. This result implies that if the threshold $\varepsilon$ for the stopping criterion in Algorithm \ref{master_algo} is made sufficiently small, the output of Algorithm \ref{master_algo} will be as close as wanted to a stationary point of \ref{master_optim}. 
		\item  \label{prop_pt6} If the equation $\mathbf a = c( \mathbf a)$ has a finite number of solutions, then the sequence $\left\lbrace  \mathbf a^s\right\rbrace $ converges to one of them.
	\end{enumerate}
	\label{cv_prop}
\end{proposition}

Proposition \ref{cv_prop} gathers all the convergence properties of Algorithm \ref{master_algo}. The three first points of Proposition \ref{cv_prop} concern 
the behavior of the sequence values $\left\lbrace f( \mathbf a^s) \right\rbrace$ of 
the objective function, whereas the three last points are about the behavior 
of the sequence $\left\lbrace  \mathbf a^s \right\rbrace$. The full proof of these 
properties is given in \cite{Tenenhaus2017}.


# The RGCCA framework 

The theoretical foundations of the Regularized Generalized Canonical Correlation 
Analysis (RGCCA) framework, previously published in \cite{Tenenhaus2011, Tenenhaus2014, Tenenhaus2015, Tenenhaus2017}, are briefly 
summarized.

## Optimization problem

A random column vector $\boldsymbol x$ of $p$ variables is assumed to exist with 
finite moments of at least order two. The random vector $\boldsymbol x$ has zero 
mean and a covariance matrix $\mathbf \Sigma$. The vector $\boldsymbol x$ is 
composed of $J$ subvectors $\boldsymbol x_j = (x_{j1}, \ldots, x_{jp_j})^\top$. 
The covariance matrix matrix $\mathbf \Sigma$ is composed of $J^2$ submatrices 
$\mathbf \Sigma_{jk} = \mathbb{E}\left[\boldsymbol x_j \boldsymbol x_k^\top\right]$. 
Let $\mathbf a_j = (a_{j1}, \ldots, a_{jp_j})^\top$ be a non-random $p_j$-dimensional 
column vector. A composite variable $y_j$ is defined as the linear combination 
of the elements of $\boldsymbol x_j$: $y_j =  \mathbf a_j^\top \boldsymbol x_j$.
Therefore the covariance between two composite variables is 
$\mathbf{a}_j^\top  \mathbf \Sigma_{jk} \mathbf{a}_k$. The RGCCA framework aims to 
extract the information shared by the $J$ random composite variables, 
taking into account an undirected graph of connections between them. 
The RGCCA framework is defined by the optimization problem (\ref{opti_theo}). 
It consists in maximizing the sum of convex functions of the covariances between 
"connected" composites $y_j$ and $y_k$ subject to specific constraints on 
the weights $\mathbf a_j$ for $j \in \{1, \ldots, J\}$.

\begin{equation}
\underset{\mathbf{a}_1, \mathbf{a}_2, \ldots,\mathbf{a}_J}{\text{max }}  f( \mathbf a_1, \ldots  \mathbf a_J) = \displaystyle  \sum_{j, k = 1}^J c_{jk} \text{
g}\left(\mathbf{a}_j^\top  \mathbf \Sigma_{jk} \mathbf{a}_k\right)\\
\text{ s.t. } \mathbf{a}_j \in \Omega_j, j =1, \ldots, J,
\label{opti_theo}
\end{equation}
where 
\begin{itemize}
\item each $\Omega_j$ is a compact set.

\item the function $g$ is any continuously differentiable convex function. 
Typical choices of $g$ are the identity (horst scheme, leading to maximizing 
the sum of covariances between block components), the absolute 
value\footnote{The scheme $g(x) = \vert x \vert$ can be included in this class 
of functions because the case $x=0$ never appears in practical applications.} 
(centroid scheme, yielding maximization of the sum of the absolute values of 
the covariances), the square function (factorial scheme, thereby maximizing the 
sum of squared covariances), or, more generally, for any even integer $m$, 
$g(x) = x^m$ (m-scheme, maximizing the power of $m$ of the sum of covariances). 
The horst scheme penalizes negative structural correlation between block 
components, while the centroid scheme and the m-scheme enable two 
components to be negatively correlated. 

\item the design matrix $\mathbf C = \lbrace c_{jk}\rbrace$ is a symmetric 
$J \times J$ matrix of non-negative elements describing the network of 
connections between blocks that the user wants to take into account. 
Usually, $c_{jk} = 1$ to two connected blocks and $0$ otherwise. 
\end{itemize}

When the diagonal of $\mathbf{C}$ is null, the convexity and the continuous 
differentiability of the function g imply that the objective function $f$ 
itself is multi-convex continuously differentiable. When at least one element 
of the diagonal of $\mathbf{C}$ is different from $0$, additional conditions have 
to be imposed on g to keep the desired property on $f$. For example, when g is 
twice differentiable, a sufficient condition is that 
$\forall x\in\mathbb{R}_+, \text{ } g'(x)\geq 0$. This condition guarantees that the 
second derivative of 
$g\left(\mathbf{a}_j^\top \mathbf{\Sigma}_{jj} \mathbf{a}_j\right)$ is positive 
definite:

\begin{equation}
\frac{\partial^2 g\left(\mathbf a_j^\top  \mathbf \Sigma_{jj} \mathbf a_j \right)}{\partial \mathbf a_j \partial \mathbf a_j^\top} = 2 \left[ g'\left( \mathbf a_j^\top  \mathbf \Sigma_{jj} \mathbf a_j \right) \mathbf \Sigma_{jj} + 2 g''\left(\mathbf a_j^\top  \mathbf \Sigma_{jj} \mathbf a_j\right)  \mathbf \Sigma_{jj}\mathbf a_j\mathbf a_j^\top \mathbf \Sigma_{jj} \right]. 
\end{equation}

All functions g considered in this paper satisfy this condition. Consequently, 
the optimization problem (\ref{opti_theo}) falls under the umbrella of the 
general optimization framework presented in Section \ref{master_algo}.

## Regularized Generalized Canonical Correlation Analysis (RGCCA) 

Several instantiations of the RGCCA framework were proposed in \cite{Tenenhaus2011, 
Tenenhaus2015, Tenenhaus2017} with 
$\Omega_j=\left\lbrace \mathbf a_j \in \mathbb{R}^{p_j}; \mathbf{a}_j^\top \mathbf M_j \mathbf{a}_j = 1 \right\rbrace$ where 
$\mathbf{M}_j$ is a symmetric positive definite matrix of order $p_j$. The optimization 
problem (\ref{opti_theo}) boils down to: 

\begin{equation}
\underset{ \mathbf a_1, \ldots  \mathbf a_J}{\text{maximize }} f( \mathbf a_1, \ldots  \mathbf a_J) = \sum_{j,k=1}^J c_{jk} \text{g}\left( \mathbf a_j^\top  \mathbf \Sigma_{jk} \mathbf{a}_k \right) \text{ s.t. }  \mathbf a_j^\top \mathbf M_j  \mathbf a_j = 1,  j=1, \ldots, J.
\label{RGCCA_optim_pop}
\end{equation}

Algorithm \ref{master_algo} can be used to solve the optimization problem 
(\ref{RGCCA_optim_pop}). This is done by updating each parameter vector, 
in turn, keeping the others fixed. Hence, we want to find an update 
$\hat{\mathbf{a}}_j\in \Omega_j=\left\lbrace \mathbf a_j \in \mathbb{R}^{p_j}; \mathbf{a}_j^\top \mathbf M_j \mathbf{a}_j = 1 \right\rbrace$ 
such that $f(\mathbf{a})\leq f(\mathbf{a}_1, \ldots, \mathbf{a}_{j-1}, \hat{\mathbf{a}}_j, \mathbf{a}_{j+1}, \ldots, \mathbf{a}_J)$. 
the RGCCA update is obtained by considering the following optimization problem:

\begin{equation}
\hat{\mathbf{a}}_j = \underset{\tilde{\mathbf{a}}_j\in\Omega_j}{\text{argmax }}  \nabla_j f(\mathbf{a})^\top \tilde{\mathbf{a}}_j = \frac{ \mathbf M_j^{-1} \mathbf \nabla_j f( \mathbf a)}{\Vert  \mathbf M_j^{-1/2} \mathbf \nabla_j f( \mathbf a) \Vert} := r_j( \mathbf a) , j=1, \ldots, J,
\label{RGCCA_update}
\end{equation}
where the partial gradient $\mathbf \nabla_j f( \mathbf a)$ of 
$f( \mathbf a)$ with respect to $\mathbf a_j$ is a $p_j$-dimensional 
column vector is given by:

\begin{equation}
 \mathbf \nabla_j f( \mathbf a)=2\sum_{k=1}^{J}c_{jk}g'\left( \mathbf a_j^\top  \mathbf \Sigma_{jk}  \mathbf a_k \right)  \mathbf \Sigma_{jk}  \mathbf a_k.
\label{grad_obj_function}
\end{equation}

A sample-based optimization problem related to (\ref{RGCCA_optim_pop}) is 
derived by considering a column partition $\mathbf X = [\mathbf X_1, \ldots, \mathbf X_j, \ldots, \mathbf X_J]$. 
In this case, each $n \times p_j$ data matrix $\mathbf X_j$ is called a block and 
represents a set of $p_j$ variables observed on $n$ individuals. 
The variables' number and nature may differ from one block to another, but the 
individuals must be the same across blocks. We assume that all variables are 
centered. The most recent formulation of RGCCA \citep{Tenenhaus2017} subsumes fifty 
years of multiblock component methods. It provides improvements to the initial 
version of RGCCA \citep{Tenenhaus2011} and is defined as the following optimization 
problem:

\begin{equation}
\underset{ \mathbf a_1, \ldots,  \mathbf a_J}{\text{maximize }}
\sum_{j, k = 1}^J c_{jk} g\left( \mathbf a_j^\top
\widehat{\mathbf{\Sigma}}_{jk} \mathbf a_k \right)\\ \text{ s.t. } \mathbf a_j^\top\widehat{\mathbf{\Sigma}}_{jj} \mathbf a_j=1, j =1, \ldots, J,
\label{opti_RGCCA_emp}
\end{equation} 

where $\widehat{\mathbf{\Sigma}}_{jk} = n^{-1} \mathbf X_j^\top  \mathbf X_k$ is an 
estimate of the inter-block covariance matrix 
$\mathbf{\Sigma}_{jk}= \mathbb{E}[\boldsymbol x_j\boldsymbol x_k^\top]$ and 
$\widehat{\mathbf{\Sigma}}_{jj}$ is an 
estimate of the intra-block covariance matrix $\mathbf{\Sigma}_{jj} = \mathbb{E}[\boldsymbol x_j\boldsymbol x_j^\top]$. In cases involving multi-collinearity within 
blocks or in high dimensional settings, one way of obtaining an estimate for the 
true covariance matrix $\mathbf{\Sigma}_{jj}$ is to consider the class of linear 
convex combinations of the identity matrix $\mathbf I$ and the sample covariance 
matrix $\mathbf{S}_{jj} = n^{-1} \mathbf X_j^\top  \mathbf X_j$. 
We then consider a version of optimization problem (\ref{opti_RGCCA_emp}) with 
$\widehat{\mathbf{\Sigma}}_{jj} = \tau_j\mathbf{I} + (1-\tau_j)\mathbf{S}_{jj}$ 
with $\tau_j \in [0,1]$ (shrinkage estimator of $\mathbf{\Sigma}_{jj}$). This 
plug-in approach leads to the RGCCA optimization problem \citep{Tenenhaus2011}. 
It is worth pointing out that for each block $j$, an appropriate shrinkage 
parameter $\tau_j$ can be obtained using various analytical formulae 
\citep[see][for instance]{Ledoit2004, Schafer2005, Chen2011}. 
As $\mathbf{M}_j$ must be positive definite, $\tau_j = 0$ can only be selected 
for a full rank data matrix $\mathbf{X}_j$.

An equivalent formulation of optimization problem (\ref{opti_RGCCA_emp}) is 
given hereafter and enables a better characterization of the objective of 
RGCCA. 
\begin{equation}
\displaystyle \underset{\mathbf{a}_1,\mathbf{a}_2, \ldots,\mathbf{a}_J}{\text{maximize }} \sum_{j, k = 1}^J c_{jk}g(\mathrm{cov}(\mathbf{X}_j\mathbf{a}_j, \mathbf{X}_k\mathbf{a}_k)) \text{ s.t. } (1-\tau_j)\mathrm{var}(\mathbf{X}_j\mathbf{a}_j) + \tau_j\Vert \mathbf{a}_j \Vert^2 = 1, j=1, \ldots,J.
\label{optim_RGCCA}
\end{equation}
Hence, the objective of RGCCA is to find block components 
$\mathbf y_j =  \mathbf X_j  \mathbf a_j, j = 1, \ldots, J$ (where $\mathbf a_j$ is a block weight 
vector of size $p_j$) summarizing the relevant information between and within 
the blocks. The $\tau_j$s are called shrinkage parameters ranging from $0$ to 
$1$ and interpolate smoothly between maximizing the covariance and maximizing 
the correlation. Setting $\tau_j$ to 0 will force the block components to 
unit variance ($\mathrm{var}(\mathbf{X}_j\mathbf{a}_j) = 1$), in which case the 
covariance criterion boils down to the correlation. Setting $\tau_j$ to 1 will 
normalize the block weight vectors ($\mathbf{a}_j^\top\mathbf{a}_j = 1$ ), which 
applies the covariance criterion. A value between $0$ and $1$ will lead to a 
compromise between the two first options and correspond to the following 
constraint 
$(1-\tau_j)\mathrm{var}(\mathbf{X}_j\mathbf{a}_j) + \tau_j \Vert \mathbf{a}_j \Vert^2 = 1$. 
We can discuss the choice of shrinkage parameters by providing interpretations 
on the properties of the resulting block components:

\begin{itemize}
\item   $\tau_j=1$ is recommended when the user wants a stable component (large 
variance) while simultaneously taking into account the correlations between 
blocks. The user must, however, be aware that variance dominates over 
correlation.

\item   $\tau_j=0$ is recommended when the user wants to maximize correlations 
between connected components. This option can yield unstable solutions in case 
of multi-collinearity and cannot be used when a data block is rank deficient 
(e.g., $n<p_j$).

\item   $0<\tau_j<1$ is a good compromise between variance and correlation: the 
block components are simultaneously stable and as well correlated as possible 
with their connected block components. This setting can be used when the data 
block is rank deficient.
\end{itemize}

In the RGCCA package, for each block, the determination of the shrinkage 
parameter can be made fully automatic by using the analytical formula proposed 
by \cite{Schafer2005} or guided by the context of an application by cross-validation 
or permutation. 
    
From optimization problem (\ref{optim_RGCCA}), the term "generalized" in the 
acronym of RGCCA embraces at least four notions. The first one relates to the 
generalization of two-block methods - including Canonical Correlation Analysis 
\citep{Hotelling1936}, Inter-battery Factor Analysis \citep{Tucker1958}, and Redundancy Analysis \citep{Wollenberg1977} - to three or more sets of variables. The second one relates to the ability to take into account some hypotheses on between-block connections: the user decides which blocks are connected and which ones are not. The third one relies on the choices of the shrinkage parameters allowing to 
capture of both correlation or covariance-based criteria. The fourth one relates 
to the function $g$ that enables considering different functions of the 
covariance. A triplet of parameters embodies this generalization: 
(g, $\tau_j, \mathbf C$) and by the fact that an arbitrary number of blocks can 
be handled. This triplet of parameters offers flexibility and allows RGCCA 
to encompass a large number of multiblock component methods that have been published 
for fifty years. Table \ref{twoblock_methods}-\ref{multiblock_hierarchical} 
gives the correspondences between the triplet (g, $\tau_j, \mathbf C$) and the 
multiblock component methods. For a complete overview, see \cite{Tenenhaus2017}. 

### Special cases 

Two families of methods have come to the fore in the field of multiblock data 
analysis. These methods rely on correlation-based or covariance-based criteria. 
Canonical correlation analysis \citep{Hotelling1936} is the seminal paper for the 
first family, and Tucker's inter-battery factor analysis \citep{Tucker1958} for 
the second one. These two methods have been extended to more than two
blocks in many ways: 

- Main contributions for generalized canonical correlation
analysis (GCCA) are found in \cite{Horst1961, Carroll1968a, Kettenring1971, 
Wold1982, Wold1985, Hanafi2007}.

- Main contributions for extending Tucker's method to more
than two blocks come from \cite{Carroll1968b, Chessel1996, Hanafi2006,
Hanafi2010, Hanafi2011, Hanafi2006, Kramer2007, Smilde2003, TenBerge1988, 
VandeGeer1984, Westerhuis1998, Wold1982, Wold1985}.

- \cite{Carroll1968b} proposed the "mixed" correlation and
covariance criterion. \cite{Wollenberg1977} combined
correlation and variance for the two-block situation (redundancy
analysis). This method is extended to the multiblock
situation in \cite{Tenenhaus2011, Tenenhaus2017}.

In the two block case, optimization problem (\ref{optim_RGCCA}) reduces to:
\begin{equation}
\underset{ \mathbf a_1,  \mathbf a_2}{\text{maximize}} \text{
cov}\left(\mathbf X_1 \mathbf a_1, \mathbf X_2 \mathbf a_2 \right) \text{ s.t. } \tau_j
\Vert  \mathbf a_j \Vert^2 + (1-\tau_j)\text{var}(\mathbf X_j \mathbf a_j) = 1, j =1,2.
\label{rCCA} 
\end{equation} 

This problem has been introduced under the name of Regularized Canonical 
Correlation Analysis \citep{Vinod1976, Leurgans1993, Shawe2004}. For various 
extreme cases $\tau_1 = 0$ or $1$ and $\tau_2 = 0$ or $1$, optimization problem 
(\ref{rCCA}) covers a situation which goes from Canonical Correlation Analysis 
\citep{Hotelling1933} to Tucker's inter-battery factor 
analysis \citep{Tucker1958}, while passing through redundancy analysis 
\citep{Wollenberg1977}. This framework corresponds exactly to the one proposed 
by \cite{Borga1997} and \cite{Burnham1996} and is reported in Table 
\ref{twoblock_methods}. 


| **Methods**                              |  $g(x)$   | $\tau_j$                  | $\mathbf{C}$                   |
|:-----------------------------------------|:---------:|:--------------------------|:------------------------------:|
| **Canonical Correlation Analysis** \citep{Hotelling1936} | $x$    | $\tau_1 = \tau_2 = 0$  | $\mathbf{C}_1 = \begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}$|
| **Inter-battery Factor Analysis** \citep{Tucker1958} or **PLS Regression** \citep{Wold1983}      | $x$    | $\tau_1 = \tau_2 = 1$  | $\mathbf{C}_1$|
| **Redundancy Analysis** \citep{Wollenberg1977}           | $x$    | $\tau_1 = 1$ ; $\tau_2 = 0$ | $\mathbf{C}_1$|
| **Regularized Redundancy Analysis** \citep{Takane2007, Bougeard2008, Qannari2005}           | $x$    | $0 \le \tau_1 \le 1$ ; $\tau_2 = 0$ | $\mathbf{C}_1$|
| **Regularized Canonical Correlation Analysis** \citep{Vinod1976, Leurgans1993, Shawe2004}           | $x$    | $0 \le \tau_1 \le 1$ ; $0 \le \tau_2 \le 1$ | $\mathbf{C}_1$|

Table: Two-block component methods. \label{twoblock_methods}

In the multiblock data analysis literature, all blocks $\mathbf X_j ,j = 1,\ldots,J$ 
are assumed to be connected, and many criteria were proposed to 
find block components satisfying some 
covariance or correlation-based optimality. Most of them are special cases of 
optimization problem (\ref{optim_RGCCA}). These multiblock component methods are 
listed in Table \ref{multiblock_methods}. PLS path modeling is also mentioned 
in this table. The great flexibility of PLS path modeling lies in the possibility 
of taking into account certain hypotheses on connections between blocks: the 
researcher decides which blocks are connected and which are not.

| **Methods**                         |  $g(x)$   | $\tau_j$                       | $\mathbf{C}$                   |
|:------------------------------------|:---------:|:-------------------------------|:------------------------------:|
| **SUMCOR** \citep{Horst1961}                             | $x$    | $\tau_j = 0, j=1, \ldots, J$ | $\mathbf{C}_2 = \begin{pmatrix} 1 & 1 & \cdots & 1 \\ 1 & 1 & \ddots & \vdots \\ \vdots & \ddots& \ddots & 1\\ 1 & \cdots & 1 & 1 \end{pmatrix}$ | 
| **SSQCOR** \citep{Kettenring1971}                        | $x^2$  | $\tau_j = 0, j=1, \ldots, J$ | $\mathbf{C}_2$ |
| **SABSCOR** \citep{Hanafi2007}                           | $|x|$  | $\tau_j = 0, j=1, \ldots, J$ | $\mathbf{C}_2$ |
| **SUMCOV-1** \citep{VandeGeer1984}                       | $x$    | $\tau_j = 1, j=1, \ldots, J$ | $\mathbf{C}_2$ |
| **SSQCOV-1** \citep{Hanafi2006}                          | $x^2$  | $\tau_j = 1, j=1, \ldots, J$ | $\mathbf{C}_2$ |
| **SABSCOV-1** \citep{Tenenhaus2011, Kramer2007}        | $|x|$  | $\tau_j = 1, j=1, \ldots, J$ | $\mathbf{C}_2$ |
| **SUMCOV-2** \citep{VandeGeer1984}                       | $x$    | $\tau_j = 1, j=1, \ldots, J$ | $\mathbf{C}_3 = \begin{pmatrix} 0 & 1 & \cdots & 1 \\ 1 & 0 & \ddots & \vdots\\ \vdots & \ddots& \ddots& 1\\ 1 & \cdots & 1 & 0 \end{pmatrix}$ |
| **SSQCOV-2** \citep{Hanafi2006}                          | $x^2$  | $\tau_j = 1, j=1, \ldots, J$ | $\mathbf{C}_3$ |
| **PLS path modeling - mode B** \citep{Wold1982, Tenenhaus2005}          | $|x|$  | $\tau_j = 0, j=1, \ldots, J$ | $c_{jk}=1$ for two connected block and $c_{jk} = 0$ otherwise |

Table: Multiblock component methods as special cases of RGCCA.\label{multiblock_methods}

\newpage

Many multiblock component methods aim to find block 
components and a global component simultaneously. For that purpose, we consider $J$ blocks, 
$\mathbf X_1, \ldots,  \mathbf X_J$ connected to a $(J + 1)$th block defined as the 
concatenation of the blocks, $\mathbf X_{J+1} = [ \mathbf X_1 ,  \mathbf X_2, \ldots,  \mathbf X_J ]$. 
Several criteria were introduced in the literature, and many are listed below.

| **Methods**                         |  $g(x)$   | $\tau_j$                       | $\mathbf{C}$                   |
|:------------------------------------|:---------:|:-------------------------------|:------------------------------:|
| **Generalized CCA** \citep{Carroll1968a}                 | $x^2$  | $\tau_j = 0, j=1, \ldots, J+1$ | $\mathbf{C}_4 = \begin{pmatrix} 0 & \cdots & 0 & 1 \\ \vdots & \ddots & \vdots & \vdots\\ 0 & \cdots & 0 & 1\\ 1 & \cdots & 1 & 0 \end{pmatrix}$ |
| **Generalized CCA** \citep{Carroll1968b}                 | $x^2$  | $\tau_j=0, j=1, \ldots, J_1$ ; $\tau_j = 1, j=J_1+1, \ldots, J$ |  $\mathbf{C}_4$ |
| **Hierarchical PCA** \citep{Wold1996}                    | $x^4$  | $\tau_j = 1, j=1, \ldots, J$ ; $\tau_{J+1} = 0$ | $\mathbf{C}_4$ |
| **Multiple Co-Inertia Analysis** \citep{Chessel1996, Westerhuis1998, Smilde2003}     | $x^2$  | $\tau_j = 1, j=1, \ldots, J$ ; $\tau_{J+1} = 0$ | $\mathbf{C}_4$ |
| **Multiple Factor Analysis** \citep{Escofier1994}     | $x^2$  | $\tau_j = 1, j=1, \ldots, J+1$ | $\mathbf{C}_4$ |
Table: Multiblock component methods in a situation of $J$ blocks: $\mathbf X_1, \ldots,  \mathbf X_J$, connected to a $(J + 1)$th block defined as the concatenation of the blocks: $\mathbf X_{J+1} = [ \mathbf X_1 ,  \mathbf X_2, \ldots,  \mathbf X_J]$.\label{multiblock_hierarchical}

The list of pre-specified multiblock component methods than can be used within 
the RGCCA package are reported below: 

```{r}
RGCCA::available_methods()
```

It is quite remarkable that the single optimization problem (\ref{optim_RGCCA}) 
offers a framework for all the multiblock component methods referenced in Table 
\ref{twoblock_methods}-\ref{multiblock_hierarchical}. From these perspectives, RGCCA provides a general 
framework for exploratory data analysis of multiblock datasets with 
immediate practical consequences for a unified statistical analysis and 
implementation strategy. The straightforward gradient-based Algorithm 
\ref{master_algo} is monotonically convergent and hits at convergence a 
stationary point. Two numerically equivalent approaches for solving 
the RGCCA optimization problem are available. A primal formulation described in 
\cite{Tenenhaus2011, Tenenhaus2017} requires the handling of matrices of dimension 
$p_j \times p_j$. A dual formulation described in \cite{Tenenhaus2015} requires 
handling matrices of dimension $n \times n$ . Therefore, the primal 
formulation of the RGCCA algorithm will be preferred when $n>p_j$, and the dual 
form will be used when $n \le p_j$.  The $\small{\texttt{rgcca()}}$ function of the RGCCA package 
implements these two formulations and automatically selects the best one.  


## Sparse Generalized Canonical Correlation Analysis

RGCCA is a component-based approach that aims to study the relationships 
between several sets of variables. The quality and interpretability of the 
RGCCA block components $\mathbf{y}_j= \mathbf{X}_j \mathbf{a}_j,j=1, \ldots,J$ 
are likely to be affected by the usefulness and relevance of 
the variables in each block.  Therefore, it is important to identify 
within each block which subsets of significant variables are active in the
relationships between blocks. For instance, biomedical data are known to be 
measurements of intrinsically parsimonious processes. 
SGCCA extends RGCCA to address this issue of variable selection \citep{Tenenhaus2014b}. 
The SGCCA optimization problem is defined as follows:

\begin{equation}
\displaystyle \underset{\mathbf{a}_1,\mathbf{a}_2, \ldots,\mathbf{a}_J}{\text{maximize }} \sum_{j, k = 1}^J c_{jk}g(\mathrm{cov}(\mathbf{X}_j\mathbf{a}_j, \mathbf{X}_k\mathbf{a}_k)) \text{ s.t. } \Vert \mathbf{a}_j \Vert_2 \le 1 \text{ and } \Vert \mathbf{a}_j \Vert_1 \le s_j, j=1,\ldots,J,
\label{optim_SGCCA}
\end{equation}

where $s_j$ is a user-defined positive constant that determines the amount of 
sparsity for $\mathbf{a}_j, j=1, \ldots,J$. The smaller the $s_j$, the larger 
the degree of sparsity for $\mathbf{a}_j$. The sparsity parameter $s_j$ is 
usually set by cross-validation or permutation procedures. Alternatively, 
values of $s_j$ can be chosen to result in desired amounts of sparsity.

SGCCA offers a sparse counterpart for all the covariance-based methods cited 
above.  

The optimization problem (\ref{optim_SGCCA}) falls into the RGCCA framework with
$\Omega_j = \lbrace  \mathbf a_j\in\mathbb{R}^{p_j}; \Vert  \mathbf a_j \Vert_2 \leq 1; \Vert  \mathbf a_j \Vert_1 \leq s_l\rbrace$. 
$\Omega_j$ is defined as the intersection between the $\ell_2$-ball of radius 
$1$ and the $\ell_1$-ball of radius $s_l \in \mathbb{R}_+^\star$ which are two 
compact sets. Hence, $\Omega_j$ is a compact set. Therefore, we can consider the 
following update for SGCCA:

\begin{equation}
	\hat{ \mathbf a}_j = \underset{\Vert \tilde{ \mathbf a}_j \Vert_2 \leq 1 ; \Vert \tilde{ \mathbf a}_j \Vert_1 \leq s_j}{\text{argmax }} \nabla_j f( \mathbf a)^\top \tilde{ \mathbf a}_j := r_j( \mathbf a).
\label{update_SGCCA}
\end{equation}

According to \cite{Witten2009a}, the solution of (\ref{update_SGCCA}) satisfies:

\begin{equation}
	r_j( \mathbf a) = \hat{ \mathbf a}_j = \frac{\mathcal{S}(\nabla_j f( \mathbf a), \lambda_j)}{\Vert \mathcal{S}(\nabla_j f( \mathbf a), \lambda_j)\Vert_2}, \text{ where } \lambda_j = \left\lbrace\begin{array}{ccc}
	0 \text{ if } & \frac{\Vert \nabla_j f( \mathbf a) \Vert_1}{\Vert \nabla_j f( \mathbf a) \Vert_2} & \leq s_j\\
	\text{find } \lambda_j \text{ such that } & \Vert \hat{ \mathbf a}_j \Vert_1 & = s_j	\end{array}\right.,
	\label{SGCCA_sol}
\end{equation}

where function $\mathcal{S}(., \lambda)$ is the soft-thresholding operator. 
When applied on a vector $\mathbf x\in\mathbb{R}^p$, this operator is defined as:

\begin{equation}
	 \mathbf u = \mathcal{S}(\mathbf x, \lambda) \Leftrightarrow u_j = \left\lbrace
	\begin{array}{ccc}
		{\mathrm{sign}}(x_j)(|x_j| -  \lambda), & \text{ if } |x_j| &> \lambda\\
		0, & \text{ if } |x_j| &\leq \lambda\\ 
	\end{array}\right., j = 1, \ldots, p.
\end{equation}

We made the assumption that the $\ell_2$-ball of radius $1$ is not included in 
the $\ell_1$-ball of radius $s_j$ and the other way round. Otherwise, 
systematically, only one of the two constraints is active. This assumption is 
true when the corresponding spheres intersect. This assumption can be translated 
into conditions on $s_j$.

The norm equivalence between $\Vert . \Vert_1$ and $\Vert . \Vert_2$ can be 
formulated as the following inequality:

\begin{equation}
	\forall \mathbf x \in \mathbb{R}^{p_j}, \text{ } \Vert \mathbf x \Vert_2 \leq \Vert \mathbf x \Vert_1 \leq \sqrt{p_j}\Vert \mathbf x \Vert_2.
\label{existence_conditions}
\end{equation}

This can be converted into a condition on $s_j$: $1 \leq s_j \leq \sqrt{p_j}$. 
When such a condition is fulfilled, the $\ell_2$-sphere of radius $1$ and the 
$\ell_1$-sphere of radius $s_j$ necessarily intersect. Within the RGCCA package, 
for consistency with the value of $\tau_j \in [0, 1]$, the level of sparsity for 
$\mathbf a_j$ is controlled with $s_j/p_j \in [1/\sqrt{p_j}, 1]$.

Several strategies, such as Binary Search or the Projection On Convex Set algorithm 
(POCS), also known as the alternating projection method \citep{Boyd2003}, can be 
used to determine the optimal $\lambda_j$ verifying the $\ell_1$-norm constraint. 
Here, a much faster approach described in \cite{Gloaguen2017} is implemented within 
the RGCCA package.

The SGCCA algorithm is similar to the RGCCA algorithm and keeps the same 
convergence properties. Empirically, we note that the S/RGCCA algorithm is 
found to be not sensitive to the starting point and usually reaches convergence 
($\small{\texttt{tol}}$= $10^{-16}$) within a few iterations.

## Higher level RGCCA algorithm

In many applications, several components per block need to be identified. 
The traditional approach consists of incorporating the single-unit RGCCA
algorithm in a deflation scheme and sequentially computing the desired number of 
components. More precisely, the RGCCA optimization problem 
returns a set of $J$ optimal block-weight vectors. denoted here 
$\mathbf a_j^{(1)}, \text{ } j = 1, \ldots, J$. Let 
$\mathbf y_j^{(1)} = \mathbf X_j \mathbf a_j^{(1)}, \text{ } j = 1, \ldots, J$ be the corresponding block 
components. Two strategies to determine higher-level weight vectors are 
presented. The first yields orthogonal block components, and the second 
yields orthogonal weight vectors. Deflation is the most straightforward way to 
add orthogonality constraints. This deflation procedure is sequential and 
consists in replacing within the RGCCA optimization problem the data matrix 
$\mathbf X_j$ by $\mathbf X_j^{(1)}$ its projection onto either: (i) the orthogonal subspace 
of $\mathbf y_j^{(1)}$ if orthogonal components are desired: 
$\mathbf X_j^{(1)} = \mathbf X_j -  \mathbf y_j^{(1)} \left( { \mathbf y_j^{(1)}}^\top  \mathbf y_j^{(1)} \right)^{-1}{ \mathbf y_j^{(1)}}^\top \mathbf X_j$, 
or (ii) the orthogonal subspace of $\mathbf a_j^{(1)}$ for orthogonal weight vectors 
$\mathbf X_j^{(1)} = \mathbf X_j - \mathbf X_j \mathbf a_j^{(1)} \left( { \mathbf a_j^{(1)}}^\top  \mathbf a_j^{(1)} \right)^{-1}{ \mathbf a_j^{(1)}}^\top$.

The second level RGCCA optimization problem boils down to:

\begin{equation}
		\underset{ \mathbf a_1, \ldots, \mathbf a_J}{\text{max }} \sum_{j, k = 1}^J c_{jk} \text{ g}\left(n^{-1} \mathbf a_j^\top {\mathbf X_j^{(1)}}^\top \mathbf X_k^{(1)}  \mathbf a_k \right)
		\text{ s.t. }  \mathbf a_j \in \Omega_j.
	\label{optim_RGCCA_orth_comp}
\end{equation} 

The optimization problem (\ref{optim_RGCCA_orth_comp}) is solved using 
Algorithm \ref{master_algo} and returns a set of optimal block-weight vectors
$\mathbf a_j^{(2)}$ and block components $\mathbf y_j^{(2)} = \mathbf X_j^{(1)} \mathbf a_j^{(2)}$, 
for $j = 1\ldots, J$. 

For orthogonal block weight vectors, 
$\mathbf y_j^{(2)} = \mathbf X_j^{(1)} \mathbf a_j^{(2)} = \mathbf X_j  \mathbf a_j^{(2)}$ 
naturally expresses as a linear combination of the original variables. 
For orthogonal block component, as $\mathbf y_j^{(1)} = \mathbf X_j \mathbf a_j^{(1)}$, 
the range space of $\mathbf X_j^{(1)}$ is included in the range space of $\mathbf X_j$. 
Therefore any block component $\mathbf y_j^{(2)}$ belonging to the range space of 
$\mathbf X_j^{(1)}$ can also be expressed in terms of the original block $\mathbf X_j$: that 
is, it exists $\mathbf a_j^{(2)^\star}$ such that $\mathbf y_j^{(2)} = \mathbf X_j^{(1)} \mathbf a_j^{(2)} = \mathbf X_j \mathbf a_j^{(2)^\star}$. 
It implies that the block components can always 
be expressed in terms of the original data matrix, whatever the deflation mode.

This deflation procedure can be iterated in a very flexible way. For instance, 
it is not necessary to keep all the blocks in the procedure at all stages: the 
number of components per block can vary from one block to another. 
This might be interesting in a supervised setting where we want to 
predict a univariate block from other blocks. In that case, the deflation 
procedure applies to all blocks except the one to predict. 

To conclude this section, when the superblock option is used, various deflation 
strategies (what to deflate and how) have been proposed in the literature.
We propose, as the default option, to deflate only the superblock with respect 
to its global components: 

$$\mathbf X_{J+1}^{(1)} = \left(\mathbf{I} -  \mathbf y_{J+1}^{(1)} \left( { \mathbf y_{J+1}^{(1)}}^\top  \mathbf y_{J+1}^{(1)} \right)^{-1}{ \mathbf y_{J+1}^{(1)}}^\top \right) \mathbf X_{J+1} = \left[ \mathbf X_1^{(1)}, \ldots, \mathbf X_J^{(1)} \right].$$

The individual blocks $\mathbf X_j^{(1)}$s are then retrieved from the deflated 
superblock. This strategy enables recovering Multiple Factor Analysis 
($\small{\texttt{ade4::mfa()/FactoMineR::MFA()}}$).
Note that, in this case, block components do not 
belong to their block space and are correlated. On the contrary, we follow the 
deflation strategy described in \cite{Chessel1996} ($\small{\texttt{ade4::mcoa()}}$) for Multiple 
Co-inertia Analysis, which is one of the most popular and established methods of 
the multiblock literature.

## Average Variance Explained

In this section, using the idea of average variance explained (AVE), the following 
indicators of model quality are defined:

- The AVE for a given block component $\mathbf{y}_j$ can be computed using the 
following formula:

\begin{equation}
\mathrm{AVE}(\mathbf X_j)=  \frac{1}{\Vert \mathbf X_j \Vert^2} \sum_{h=1}^{p_j} \text{var}(\mathbf{x}_{jh}) \times \text{cor}^2(\mathbf{x}_{jh},\mathbf{y}_j).
\label{AVE_X}
\end{equation}

- A global indicator of model quality can be obtained by considering  a weighted 
sum of these individual AVEs. This outer AVE is defined as:

\begin{equation}
\displaystyle \mathrm{AVE(outer model)} = \left( 1/\sum_j \Vert \mathbf X_j \Vert^2 \right) \sum_j \Vert \mathbf X_j \Vert^2 \mathrm{AVE}(\mathbf X_j).
\label{AVE_outer}
\end{equation}

However, the previous quantities do not take into account the correlations
between blocks. Therefore, another indicator of model quality is the inner
AVE, defined as follows:

\begin{equation}
\displaystyle \mathrm{AVE(inner model)} = \left( 1/\sum_{j<k} c_{jk} \right) \sum_{j<k} c_{jk} \mathrm{cor}^2(\mathbf{y}_j , \mathbf{y}_k).
\label{AVE_inner}
\end{equation}

All these quantities vary between 0 and 1 and reflect important properties of
the model.  

Equation (\ref{AVE_X}) is defined for a specific block component. The cumulative AVE is 
obtained by summing these individual AVEs over the different components. 
However, this sum applies only to orthogonal block components.  For correlated 
components, we follow the QR-orthogonalization procedure described in \cite{Zou2006} 
to consider only the increase of AVE due to adding the new 
components. 

Guidelines describing R/SGCCA in practice are provided in \cite{Garali2018}.
The usefulness and versatility of the RGCCA package are illustrated in the next 
section. 

# Practical session

We now present the functions implemented in the RGCCA package and give examples
of how they can be used. The list of functions can be found in Table 
\ref{exported_functions}.

| Function                |  Description                   |
|:------------------------|:-------------------------------|
| $\small{\texttt{rgcca}}$                 | Main entry point of the package, this function allows fitting a R/SGCCA model on a multiblock dataset.           |
| $\small{\texttt{rgcca\_transform}}$       | Use a fitted R/SGCCA model to compute the block components of unseen individuals.                                |
| $\small{\texttt{rgcca\_predict}}$         | Train a caret model on the block components of a fitted R/SGCCA model and predict values for unseen individuals. |
| $\small{\texttt{rgcca\_cv}}$              | Find the best set of parameters for a R/SGCCA model using cross-validation.                                      |
| $\small{\texttt{rgcca\_permutation}}$     | Find the best set of parameters for a R/SGCCA model using a permutation strategy.                                |
| $\small{\texttt{rgcca\_bootstrap}}$       | Evaluate the significance of the block-weight vectors produced by a R/SGCCA model using bootstrap.               |
| $\small{\texttt{rgcca\_stability}}$       | Select the most stable variables of a R/SGCCA model using their VIPs.                                            |
| $\small{\texttt{print/plot}}$            | Print and plot methods for outputs of functions $\small{\texttt{rgcca}}$, $\small{\texttt{rgcca\_cv}}$, $\small{\texttt{rgcca\_permutation}}$, $\small{\texttt{rgcca\_bootstrap}}$, and $\small{\texttt{rgcca\_stability}}$. |
Table: Functions implemented in the RGCCA package. \label{exported_functions}

## RGCCA for the Russett dataset.

In this section, we reproduce some of the results presented in 
\cite{Tenenhaus2011} from the Russett data. The Russett dataset is available within 
the RGCCA package. The Russett dataset \citep{Russett1964} is studied in \cite{Gifi1990}. 
Russett collected this data to study relationships between Agricultural 
Inequality, Industrial Development, and Political Instability.

```{r}
library(RGCCA)
data(Russett)
colnames(Russett)
```

The first step of the analysis is to define the blocks. Three blocks of 
variables have been defined for 47 countries. The variables that compose each 
block have been defined according to the nature of the variables.

-   The first block $\mathbf{X}_1$ = [$\small{\texttt{gini}}$, $\small{\texttt{farm}}$, $\small{\texttt{rent}}$] is related to 
"Agricultural Inequality":
    -   $\small{\texttt{gini}}$ = Inequality of land distribution,
    -   $\small{\texttt{farm}}$ = \% farmers that own half of the land (\> 50),
    -   $\small{\texttt{rent}}$ = \% farmers that rent all their land.
-   The second block $\mathbf{X}_2$ = [$\small{\texttt{gnpr}}$, $\small{\texttt{labo}}$] describes "Industrial Development":
    -   $\small{\texttt{gnpr}}$ = Gross national product per capita (\$1955),
    -   $\small{\texttt{labo}}$ = `\% of labor force employed in agriculture.
-   The third one $\mathbf{X}_3$ = [$\small{\texttt{inst}}$, $\small{\texttt{ecks}}$, $\small{\texttt{death}}$] measures "Political Instability":
    -   $\small{\texttt{inst}}$ = Instability of executive (45-61),
    -   $\small{\texttt{ecks}}$ = Number of violent internal war incidents (46-61),
    -   $\small{\texttt{death}}$ = Number of people killed as a result of civic group violence (50-62).
    -   $\small{\texttt{demo}}$ = Political regime: stable democracy, unstable democracy or 
    dictatorship. Due to redundancy, the dummy variable “unstable democracy” 
    has been left out.

The different blocks of variables $\mathbf X_1, \ldots, \mathbf X_J$ are arranged in the list format.

```{r}
A <- list(
  Agric = Russett[, c("gini", "farm", "rent")],
  Ind = Russett[, c("gnpr", "labo")],
  Polit = Russett[, c("inst", "ecks",  "death", "demostab", "dictator")])

lab <- factor(
  apply(Russett[, 9:11], 1, which.max),
  labels = c("demost", "demoinst", "dict")
)
```

**Preprocessing.** In general, and especially for the covariance-based 
criterion, the data blocks might be preprocessed to ensure comparability 
between variables and blocks. In order to ensure comparability between 
variables, standardization is applied (zero mean and unit variance). Such 
preprocessing is reached by setting the $\small{\texttt{scale}}$ argument to $\small{\texttt{TRUE}}$ (default 
value) in the $\small{\texttt{rgcca()}}$ function. To make blocks comparable, a possible 
strategy is to standardize the variables and divide each block by 
the square root of its number of variables \citep{Westerhuis1998}. This two-step 
procedure leads to $\mathrm{tr}(\mathbf X_j^\top \mathbf X_j )=n$ for each block (i.e. the 
sum of the eigenvalues of the covariance matrix of $\mathbf X_j$ is equal to $1$ 
whatever the block). Such a preprocessing is reached by 
setting the $\small \texttt{scale\_block}$ argument to 
$\small \texttt{TRUE}$ or $\small{\texttt{"inertia"}}$ (default value) in the 
$\small{\texttt{rgcca()}}$ function. 
If  $\small{\texttt{scale\_block = "lambda1"}}$, each block is divided by the 
square root of the highest eigenvalue of its empirical covariance matrix.  If 
standardization is applied ($\small{\texttt{scale = TRUE}}$), the block scaling is applied on 
the result of the standardization.

**Definition of the design matrix** $\mathbf{C}$. From Russett's hypotheses, it 
is difficult for a country to escape dictatorship when agricultural 
inequality is above average, and industrial development is below average. These 
hypotheses on the relationships between blocks are encoded through the design 
matrix $\mathbf{C}$; usually $c_{jk} = 1$ for two connected blocks and $0$ 
otherwise. Therefore, we have decided to connect Agricultural Inequality to 
Political Instability ($c_{13} = 1$), Industrial Development to Political 
Instability ($c_{23} = 1$) and to not connect Agricultural Inequality to 
Industrial Development ($c_{12} = 0$). The resulting design matrix $\mathbf{C}$ 
is:

```{r}
#Define the design matrix C.
C <- matrix(c(0, 0, 1,
              0, 0, 1,
              1, 1, 0), 3, 3)

C
```

**Choice of the scheme function g**. Typical choices of scheme functions are $g(x) = x, x^2$, or $\vert x \vert$.  According to \cite{VandeGeer1984}, a fair model is a model where all blocks contribute equally to the solution in opposition to a model dominated by only a few of the $J$ sets. If fairness is a major objective, the user must choose $m=1$. $m>1$ is preferable if the user wants to discriminate between blocks. In practice, $m$ is equal to $1$, $2$ or $4$. The higher the value of $m$, the more the method acts as block selector \citep{Tenenhaus2017}.

RGCCA using the pre-defined design matrix $\mathbf{C}$, 
the factorial scheme ($g(x) = x^2$), $\tau = 1$ for all blocks (full covariance criterion) 
and a number of (orthogonal) components equal to $2$ for all blocks is obtained by specifying 
appropriately  the arguments $\small{\texttt{connection}}$, $\small{\texttt{scheme}}$, $\small{\texttt{tau}}$, $\small{\texttt{ncomp}}$, $\small{\texttt{comp\_orth}}$ in 
$\small{\texttt{rgcca()}}$. $\small{\texttt{verbose}}$ (default value = $\small{\texttt{TRUE}}$) indicates that the progress will be reported 
while computing and that a plot illustrating the convergence of the algorithm will be displayed.

```{r}
fit <- rgcca(blocks = A, connection = C,
             tau = 1, ncomp = 2,
             scheme = "factorial",
             scale = TRUE,
             scale_block = FALSE,
             comp_orth = TRUE,
             verbose = FALSE)
```

The $\small{\texttt{summary()}}$ function allows summarizing the RGCCA analysis.

```{r}
summary(fit)
```

The block-weight vectors solution of the optimization problem (\ref{optim_RGCCA}) 
are available as output of the $\small{\texttt{rgcca()}}$ function in $\small{\texttt{fit\$a}}$ and correspond exactly to the weight vectors reported in Figure 5 of \cite{Tenenhaus2011}. 
It is possible to display specific block-weight vector(s) ($\small{\texttt{type = "weight"}}$) block-loadings vector(s) ($\small{\texttt{type = "loadings"}}$) using the generic $\small{\texttt{plot()}}$ function and specifying the arguments $\small{\texttt{block}}$ and $\small{\texttt{comp}}$ accordingly. The
${ \mathbf a_j^{(k)}}^\star$, mentioned in Section [Higher level RGCCA algorithm], 
are available in $\small{\texttt{fit\$astar}}$.

```{r, fig.height = 12, fig.width=18, fig.cap = 'Block-weight vectors of a fitted RGCCA model.', fig.pos = "H"}
plot(fit, type = "weight", block = 1:3, comp = 1,
     display_order = FALSE, cex = 2)
```

As a component-based method, the RGCCA package provides block components as output 
of the $\small{\texttt{rgcca()}}$ function in $\small{\texttt{fit\$Y}}$ and graphical representations, including 
factor plot ($\small{\texttt{type = "sample"}}$), correlation circle ($\small{\texttt{type = "cor\_circle"}}$) or 
biplot ($\small{\texttt{type = "biplot"}}$).  This graphical display allows visualizing the sources of variability within blocks, 
the relationships between variables within and between blocks, and the amount of correlation between blocks. 
The graphical display of the countries obtained by crossing $\mathbf X_1  \mathbf a_1$ = Agricultural Inequality 
and $\mathbf X_2  \mathbf a_2$ = Industrial Development and marked with their political regime in 1960 is shown below.

```{r, fig.align='center', fig.cap = '\\label{fig:sample}Graphical display of the countries by drawing the block component of the first block against the block component of the second block, colored according to their political regime.', fig.height = 12, fig.width=18, fig.pos = "H"}
plot(fit, type = "sample",
     block = 1:2, comp = 1,
     resp = lab, repel = TRUE, cex = 2)
```

Countries aggregate together when they share similarities. It may be noted that 
the lower right quadrant concentrates on dictatorships. It is difficult for a 
country to escape dictatorship when its industrial development is below average, 
and its agricultural inequality is above average. It is worth pointing out that 
some unstable democracies located in this quadrant (or close to it) became 
dictatorships for a period of time after 1960: Greece (1967-1974), Brazil 
(1964-1985), Chili (1973-1990), and Argentina (1966-1973). 

The AVEs of the different blocks are reported in the axes 
of Figure \ref{fig:sample}.
All AVEs (defined in \ref{AVE_X}-\ref{AVE_inner}) are available as output of 
the $\small{\texttt{rgcca()}}$ function in $\small{\texttt{fit\$AVE}}$. These indicators of model quality can 
also be visualized using the generic $\small{\texttt{plot()}}$ function.

```{r, fig.align='center', fig.cap = 'Average Variance Explained of the different blocks.', fig.height = 8, fig.width=18, fig.pos = "H"}
plot(fit, type = "ave", cex = 2)
```

The strength of the relations between each block component and each variable 
can be visualized using correlation circles or biplot representations.

```{r, fig.align='center', fig.cap = 'Correlation circle associated with the first two components of the Agriculture block.', fig.height = 12, fig.width=18, fig.pos = "H"}
plot(fit, type = "cor_circle", block = 1, comp = 1:2, 
     display_blocks = 1:3, cex = 2)
```

By default, all the variables are displayed on the correlation circle. However, 
it is possible to choose the block(s) to display ($\small{\texttt{display\_blocks}}$) in the 
correlation_circle. 

```{r, fig.align='center', fig.cap = 'Biplot associated with the first two components ofthe Agriculture block.', fig.height = 12, fig.width=18, fig.pos = "H"}
plot(fit, type = "biplot", block = 1, 
     comp = 1:2, repel = TRUE, 
     resp = lab, cex = 2,
     show_arrow = TRUE)
```

As we will see in the next section, when the superblock option is considered 
($\small{\texttt{superblock = TRUE}}$ or $\small{\texttt{method}}$ set to a method that induces the use of 
superblock), global components can be derived. The space spanned by the global 
components can be viewed as a consensus space that integrates all the 
modalities and facilitates the visualization of the results and their 
interpretation.

**Assessment of the reliability of parameter estimates.** It is possible to use 
a bootstrap resampling method to assess the reliability of parameter estimates (block-weight/loading vectors) obtained using RGCCA. $\small{\texttt{B = n\_boot}}$ bootstrap 
samples of the same size as the original data are repeatedly sampled with 
replacement from the original data. RGCCA is then applied to each bootstrap 
sample to obtain the RGCCA estimates. We calculate the standard deviation of 
the estimates across the bootstrap samples, from which we derive bootstrap 
confidence intervals, t-ratio (defined as the ratio of the parameter estimate 
to its bootstrap estimate of the standard deviation), and p-value (the p-value 
is computed by assuming that the ratio of the parameter estimate to its standard 
deviation follows the standardized normal distribution), to indicate how reliably 
parameters were estimated. Since several p-values are constructed simultaneously, 
FDR correction can be applied to control the False Discovery Rate. This 
function is available using the $\small{\texttt{rgcca\_bootstrap()}}$ function of the RGCCA 
package.

```{r, cache = TRUE, message = FALSE}
boot_out <- rgcca_bootstrap(fit, n_boot = 500, n_cores = 1)
```

The bootstrap results are detailed using the $\small{\texttt{summary()}}$ function,

```{r, size = "tiny"}
summary(boot_out, block = 1:3, ncomp = 1)
```

and displayed using the $\small{\texttt{plot()}}$ function.


```{r, fig.cap = 'Bootstrap confidence intervals for the block-weight vectors.', fig.height = 12, fig.width=18, fig.pos = "H"}
plot(boot_out, type = "weight", 
     block = 1:3, comp = 1, 
     display_order = FALSE, cex = 2,
     show_stars = TRUE)
```

Each weight is shown along with its associated bootstrap confidence interval 
and stars ($\small{\texttt{show\_stars = TRUE}}$) reflecting the p-value of assigning a strictly positive or negative weight to this variable.

## RGCCA with superblock

In this section, we consider Multiple Co-Inertia Analysis \citep{Chessel1996} 
\citep[MCOA, also called MCIA in][]{Cantini2021} with $2$ components per block. 

See $\small{\texttt{available\_methods()}}$ for a list of pre-specified 
multiblock component methods. 

```{r}
fit.mcoa <- rgcca(blocks = A, method = "mcoa", ncomp = 2)
```

Interestingly, the $\small{\texttt{summary()}}$ function reports the arguments
implicitly specified to perform MCOA. 

```{r}
summary(fit.mcoa)
```

It is possible to display specific output as previously using the generic 
$\small{\texttt{plot()}}$ function by specifying the argument $\small{\texttt{type}}$ accordingly. MCOA enables 
individuals to be represented in the space spanned by the first global 
components. The biplot representation associated with this consensus space 
is given below.


```{r, fig.align='center', fig.cap = 'Biplot of the countries obtained by crossing the two first components of the superblock. Individuals are colored according to their political regime and variables according to their block membership.', fig.height = 12, fig.width=18, fig.pos = "H"}
plot(fit.mcoa, type = "biplot", 
     block = 4, comp = 1:2, 
     response = lab, 
     repel = TRUE, cex = 2)
```

As previously, this model can be easily bootstrapped using the 
$\small{\texttt{rgcca\_bootstrap()}}$ function, and the bootstrap confidence intervals are still 
available using the $\small{\texttt{summary()}}$ and $\small{\texttt{plot()}}$ functions.

## Choice of the shrinkage parameters

Three fully automatic strategies are proposed to select the optimal shrinkage 
parameters:

**The Schafer and Strimmer analytical formula.** For each block $j$, an "optimal" 
shrinkage parameter $\tau_j$ can be obtained using the Schafer and Strimmer 
analytical formula \citep{Schafer2005} by setting the $\small{\texttt{tau}}$ argument of the $\small{\texttt{rgcca()}}$ 
function to $\small{\texttt{"optimal"}}$.

```{r}
fit <- rgcca(blocks = A, connection = C,
             tau = "optimal", scheme = "factorial")
```

The optimal shrinkage parameters are given by:

```{r}
fit$call$tau
```

This automatic estimation of the shrinkage parameters allows one to come closer 
to the correlation criterion, even in the case of high multicollinearity or 
when the number of individuals is smaller than the number of variables.

As previously, all the fitted RGCCA objects can be visualized/bootstrapped using 
the $\small{\texttt{summary()}}$, $\small{\texttt{plot()}}$ and $\small{\texttt{rgcca\_bootstrap()}}$ functions.

**Permutation strategy.** A permutation-based strategy very similar to the one 
proposed in \cite{Witten2009a} has also been integrated within the RGCCA package 
through the $\small{\texttt{rgcca\_permutation()}}$ function. This function is used to 
select the regularization parameters for R/SGCCA automatically.

For each set of regularization parameters (generally this will be a 
$J$-dimensional vector), the following steps are performed:

- S/RGCCA is run on the original data 
$\mathbf X_1, \ldots, \mathbf X_J$, and we record the value of the objective 
function, denoted $t$.

- $\small{\texttt{n\_perm}}$ times, the rows of $\mathbf X_1, \ldots, \mathbf X_J$ are
randomly permuted to obtained permuted data sets $\mathbf X_1^*, \ldots, \mathbf X_J^*$.
S/RGCCA is then run on these permuted data sets, and we record the value of the objective
function, denoted $t^*$.

- The resulting p-value is given by the fraction of permuted
$t^*$ that exceeds the $t$ obtained from the non-permuted blocks.

- The resulting zstat is defined as $\frac{t-\text{mean}(t^*)}{\text{sd}(t^*)}$.

The best set of tuning parameters is then the set that yields the highest zstat.
This procedure is available through the $\small{\texttt{rgcca\_permutation}}$ function.

```{r}
set.seed(123)
perm_out <- rgcca_permutation(blocks = A, connection = C,
                              par_type = "tau",
                              par_length = 10,
                              n_cores = 1,
                              n_perms = 10)
```

By default, the $\small{\texttt{rgcca\_permutation}}$ function generates 
10 sets of tuning parameters uniformly between some minimal values (0 for 
RGCCA and $1/\text{sqrt}(\text{ncol})$ for SGCCA) and 1. 
Results of the permutation procedure are summarized using the generic 
$\small{\texttt{summary()}}$ function,

```{r, width=30}
summary(perm_out)
```

and displayed using the generic $\small{\texttt{plot()}}$ function.

```{r, fig.height = 12, fig.width=18, fig.pos = "H", fig.cap = "Values of the objective function of RGCCA against the sets of tuning parameters, triangles correspond to evaluations on non-permuted datasets."}
plot(perm_out, cex = 2)
```

The fitted permutation object, $\small{\texttt{perm\_out}}$, can be directly provided as the output of $\small{\texttt{rgcca()}}$ and visualized/bootstrapped as usual.

```{r}
fit <- rgcca(perm_out)
```

Of course, it is  possible to define explicitly the combination of 
regularization parameters to be tested. In that case, a matrix of dimension 
$K \times J$ is required. Each row of this matrix corresponds to one set of 
tuning parameters. Alternatively, a numeric vector of length $J$ indicating 
the maximum range values to be tested can be given. The set of parameters 
is then uniformly generated between the minimum values (0 for RGCCA and 
$1/\text{sqrt}(\text{ncol})$ for SGCCA) and the maximum values specified by 
the user with $\small{\texttt{par\_value}}$.

**Cross-validation strategy.** The optimal tuning parameters can also be 
obtained by cross-validation. We will illustrate this in the next
section in the context of SGCCA.

# High dimensional case study: Glioma Data

**Biological problem.** Brain tumors are children's most common solid tumors 
and have the highest mortality rate of all pediatric cancers. Despite 
advances in multimodality therapy, children with pHGG invariably have an overall 
survival of around 20% at 5 years. Depending on their location (e.g. brainstem, 
central nuclei, or supratentorial), pHGG present different characteristics in 
terms of radiological appearance, histology, and prognosis. Our hypothesis is 
that pHGG have different genetic origins and oncogenic pathways depending on 
their location. Thus, the biological processes involved in the development of 
the tumor may be different from one location to another, as has been 
frequently suggested.

**Description of the data.** Pretreatment frozen tumor samples were obtained 
from 53 children with newly diagnosed pHGG from Necker Enfants Malades (Paris, France)
\citep{Puget2012}. The 53 tumors are divided into 3 locations: supratentorial 
(HEMI), central nuclei (MIDL), and brain stem (DIPG). The final dataset is 
organized into 3 blocks of variables defined for the 53 tumors: the first block 
$\mathbf{X}_1$ provides the expression of $15702$ genes (GE). The second block 
$\mathbf{X}_2$ contains the imbalances of $1229$ segments (CGH) of chromosomes. 
$\mathbf{X}_3$ is a block of dummy variables describing the categorical variable 
location. One dummy variable has been left out because of redundancy with the 
others.

The next lines of code can be run to download the dataset:

```{r eval = FALSE}
# Download the dataset's package at http://biodev.cea.fr/sgcca/.
# --> gliomaData_0.4.tar.gz
if (!("gliomaData" %in% rownames(installed.packages()))) {
  destfile <- tempfile()
  download.file("http://biodev.cea.fr/sgcca/gliomaData_0.4.tar.gz", destfile)
  install.packages(destfile, repos = NULL, type = "source")
}
```

```{r include = FALSE}
knitr::opts_chunk$set(eval = "gliomaData" %in% rownames(installed.packages()))
```

```{r}
data(ge_cgh_locIGR, package = "gliomaData")

blocks <- ge_cgh_locIGR$multiblocks
Loc <- factor(ge_cgh_locIGR$y)
levels(Loc) <- colnames(ge_cgh_locIGR$multiblocks$y)
blocks[[3]] <- Loc

# check dimensions of the blocks
vapply(blocks, NCOL, FUN.VALUE = 1L)
```

We impose $\mathbf{X}_1$ and $\mathbf{X}_2$ to be connected to $\mathbf{X}_3$. 
This design is commonly used in many applications and is oriented toward 
predicting the location. The argument $\small{\texttt{response = 3}}$ of the $\small{\texttt{rgcca()}}$ function 
encodes this design. 

```{r}
fit.rgcca <- rgcca(blocks = blocks, response = 3, ncomp = 2, verbose = FALSE)
```

When the response variable is qualitative, two steps are implicitly performed: 
(i) disjunctive coding and (ii)  the associated shrinkage parameter is set to 
$0$ regardless of the value specified by the user. 

```{r}
fit.rgcca$call$connection
fit.rgcca$call$tau
```

From the dimension of each block ($n>p$ or $n\leq p$), $\small{\texttt{rgcca()}}$ 
selects automatically the dual formulation for $\mathbf{X}_1$ and 
$\mathbf{X}_2$ and the primal one for $\mathbf{X}_3$.
The formulation used for each block is returned using the following command:
  
```{r}
fit.rgcca$primal_dual
```

The dual formulation makes the RGCCA algorithm highly efficient, even in a 
high-dimensional setting.

```{r}
system.time(
  rgcca(blocks = blocks, response = 3)
)
```

RGCCA enables visual inspection of the spatial relationships between classes. 
This facilitates assessment of the quality of the classification and makes it 
possible to determine which components capture the discriminant 
information readily.

```{r, fig.align='center', fig.height = 12, fig.width=18, fig.pos = "H", fig.cap = "Graphical display of the tumors obtained by crossing the block components, and colored according to their location."}
plot(fit.rgcca, type = "sample", block = 1:2,
     comp = 1, response = Loc, cex = 2)
```

For easier interpretation of the results, especially in high-dimensional 
settings, it is often appropriate to add penalties promoting sparsity within 
the RGCCA optimization problem. For that purpose, an $\ell_1$ penalization on 
the weight vectors $\mathbf{a}_1, \ldots, \mathbf{a}_J$ is applied. the 
$\small{\texttt{sparsity}}$ argument of $\small{\texttt{rgcca()}}$ varies between 1/sqrt(ncol) and 1 (larger 
values of $\small{\texttt{sparsity}}$ correspond to less penalization) and controls the amount of 
sparsity of the weight vectors $\mathbf{a}_1, \ldots, \mathbf{a}_J$. If 
$\small{\texttt{sparsity}}$ is a vector, $\ell_1$-penalties are the same for all the weights 
corresponding to the same block but different components:
  
  ```{=tex}
\begin{equation}
\forall h, \Vert \mathbf{a}_j^{(h)} \Vert_{\ell_1} \leq \small{\texttt{sparsity}}_j \sqrt{p_j},
\end{equation}
```
with $p_j$ the number of variables of $\mathbf X_j$.

If $\small{\texttt{sparsity}}$ is a matrix, row $h$ of $\small{\texttt{sparsity}}$ defines the constraints 
applied to the weights corresponding to components $h$:
  
  ```{=tex}
\begin{equation}
\forall h, \Vert \mathbf{a}_j^{(h)} \Vert_{\ell_1} \leq \small{\texttt{sparsity}}_{h,j} \sqrt{p_j}.
\end{equation}
```

## SGCCA for the Glioma dataset

The algorithm associated with the optimization problem 
(\ref{optim_SGCCA}) is available through the function $\small{\texttt{rgcca()}}$ with the 
argument $\small{\texttt{method = "sgcca"}}$. 

```{r}
fit.sgcca <- rgcca(blocks = blocks, response = 3, ncomp = 2,
                   sparsity = c(0.0710, 0.2000, 1),
                   verbose = FALSE)
```

The $\small{\texttt{summary()}}$ function allows summarizing the SGCCA analysis,

```{r}
summary(fit.sgcca)
```

and the $\small{\texttt{plot()}}$ returns the same graphical displays as RGCCA. 
We skip these representations for sake of brevity.

Of course, it is still possible to determine the optimal sparsity parameters by permutation. 
This is made possible by setting the $\small{\texttt{par\_type}}$ argument to $\small{\texttt{"sparsity"}}$ (instead of 
$\small{\texttt{"tau"}}$) within the $\small{\texttt{rgcca\_permutation()}}$ function. However, we will use another approach
in this section.

**Cross-validation strategy.** The 
optimal tuning parameters can be determined by cross-validating 
different indicators of quality, namely:

- For classification: $\small{\texttt{Accuracy}}$, $\small{\texttt{Kappa}}$, $\small{\texttt{F1}}$, $\small{\texttt{Sensitivity}}$, $\small{\texttt{Specificity}}$, 
$\small{\texttt{Pos\_Pred\_Value}}$, $\small{\texttt{Neg\_Pred\_Value}}$, $\small{\texttt{Precision}}$, $\small{\texttt{Recall}}$, $\small{\texttt{Detection\_Rate}}$, and
$\small{\texttt{Balanced\_Accuracy}}$. 

- For regression: $\small{\texttt{RMSE}}$ and $\small{\texttt{MAE}}$.

This cross-validation protocol is made available through the 
$\small{\texttt{rgcca\_cv}}$ function and is used here for predicting the 
location of the tumors.

In this situation, the goal is to maximize the 
cross-validated  accuracy ($\small{\texttt{metric = "Accuracy"}}$) in a model where we try to 
predict the response block from all the block components with a user-defined 
classifier ($\small{\texttt{prediction\_model = "lda"}}$). Also, we decide to upper bound the sparsity 
parameters for $\mathbf X_1$ and $\mathbf X_2$ to $0.2$ to achieve an 
attractive amount of sparsity. 

```{r, cache = TRUE}
set.seed(27) #my favorite number
inTraining <- caret::createDataPartition(
  blocks[[3]], p = .75, list = FALSE
)
training <- lapply(blocks, function(x) as.matrix(x)[inTraining, , drop = FALSE])
testing <- lapply(blocks, function(x) as.matrix(x)[-inTraining, , drop = FALSE])

cv_out <- rgcca_cv(blocks = training, response = 3,
                   par_type = "sparsity",
                   par_value = c(.2, .2, 0),
                   par_length = 10,
                   prediction_model = "lda",
                   validation = "kfold",
                   k = 3, n_run = 5, metric = "Accuracy",
                   n_cores = 2)
```

$\small{\texttt{rgcca\_cv()}}$ relies on the $\small{\texttt{caret}}$ package. 
As a direct consequence, an astonishingly large number of models are made 
available (see $\small{\texttt{caret::modelLookup()}}$). Results of the 
cross-validation procedure are reported using the generic 
$\small{\texttt{summary()}}$ function,

```{r}
summary(cv_out)
```

and displayed using the generic $\small{\texttt{plot()}}$ function.

```{r, fig.height = 12, fig.width=18, fig.pos = "H", fig.cap = "Accuracies of the models on the different validation folds against the sets of tuning parameters."}
plot(cv_out, cex = 2)
```

As previously, the optimal sparsity parameters can be used to fit a new model,
and the resulting optimal model can be visualized/bootstrapped.

```{r}
fit <- rgcca(cv_out)
summary(fit)
```

Note that the sparsity parameter associated with $\mathbf{X}_3$ switches 
automatically to $\tau_3 = 0$. This choice is 
justified by the fact that we were not looking for a block component 
$\mathbf y_3$ that explained its own block well (since $\mathbf{X}_3$ is a group 
coding matrix) but one that is correlated with its neighboring components. 

At last, $\small{\texttt{rgcca\_predict()}}$ can be used for predicting new blocks,

```{r}
pred <- rgcca_predict(fit, blocks_test = testing, prediction_model = "lda")
```

and a $\small{\texttt{caret}}$ summary of the performances can be reported.

```{r}
pred$confusion$test
```

If, for a specific reason, only the block components are wanted for the test 
set, the function $\small{\texttt{rgcca\_transform}}$ can be used.

```{r}
projection <- rgcca_transform(fit, blocks_test = testing)
```

**Stability procedure.** It is possible to stabilize the selected variables using 
the following procedure.

\cite{Tenenhaus1998} defines the Variable Importance in Projection
(VIP) score for the PLS method. This score is used for variable selection: 
the higher the score, the more important the variable. We use 
this idea to propose a procedure for evaluating the stability of the variable
selection procedure of SGCCA. This procedure relies on the following 
score:

\begin{equation}
\displaystyle \mathrm{VIP}(\mathbf{x}_{jh}) = \frac{1}{K} \sum_{k=1}^K \left(\mathbf{a}_{jh}^{(k)2} \mathrm{AVE}\left(\mathbf X_j^{(k)}\right)\right).
\label{VIP}
\end{equation}

SGCCA is run several times using a bootstrap resampling procedure. For each model,
the VIPs are computed, and the variables with the higher VIPs averaged over the
different models are kept. This procedure is available  through the 
$\small{\texttt{rgcca\_stability}}$ function.

```{r, cache = TRUE, message = FALSE}
fit_stab <- rgcca_stability(fit,
                            keep = vapply(
                              fit$a, function(x) mean(x != 0),
                              FUN.VALUE = 1.0
                            ),
                            n_boot = 100, verbose = TRUE, n_cores = 2)
```

Once the most stable variables have been found, a new model using these 
variables is automatically fitted. This last model can be visualized using
the usual $\small{\texttt{summary()}}$ and $\small{\texttt{plot()}}$ functions.
  
```{r, fig.height = 12, fig.width=18, fig.align='center', fig.cap = 'Graphical display of the tumors obtained by crossing the components of GE1 and CGH1, and colored according to their location.', fig.pos = "H"}
plot(fit_stab, type = "sample", block = 1:2,
     comp = 1, resp = as.character(Loc)[inTraining],
     cex = 2
     )
```

We can finally apply the bootstrap procedure on the most stable variables.

```{r, cache = TRUE}
boot_out <- rgcca_bootstrap(fit_stab, n_boot = 500)
```
  
The bootstrap results can be visualized using the generic $\small{\texttt{plot()}}$ function.
We use the $\small{\texttt{n\_mark}}$ parameter to display the top 50 variables of GE.


```{r, fig.height = 12, fig.width=18, fig.pos = "H", fig.cap = "Bootstrap confidence intervals for the block-weight vectors associated to block GE."}
plot(boot_out, block = 1,
     display_order = FALSE,
     n_mark = 50, cex = 1.5, cex_sub = 17,
     show_star = TRUE)
```


# Conclusion

The RGCCA framework gathers fifty years of multiblock component methods and
offers a unified implementation strategy for these methods. The RGCCA package 
is available on the Comprehensive R Archive Network (CRAN) and GitHub 
https://github.com/rgcca-factory/RGCCA. This release of the RGCCA package 
includes:

\begin{itemize}
\item Several strategies for determining the shrinkage 
parameters/level of sparsity automatically: Schaffer \& Strimmer's analytical formulae, 
cross-validation, or permutation strategy.

\item A bootstrap resampling procedure for assessing the reliability of the 
parameter estimates of S/RGCCA.

\item Dedicated functions for graphical displays of the output of RGCCA 
(sample plot, correlation circle, biplot, ...).

\item Various implementation strategies for orthogonal block-components or 
orthogonal block-weight vectors.


\item Strategies for handling missing data. Specifically, multiblock data 
faces two types of missing data structure: (i) if an 
observation $i$ has missing values on a whole block j and (ii) if an 
observation i has some missing values on a block j (but not all). For these two 
situations, we exploit the algorithmic solution proposed for PLS 
path modeling to deal with missing data \citep[see][]{Tenenhaus2005}.

\item Special attention has been paid to providing a bunch of "mathematical" unit tests 
which, in a sense, guarantee the implementation quality. Also, when appropriate, a particular focus was given to recovering the results of other R packages of the literature , including $\small{\texttt{ade4}}$ and $\small{\texttt{FactoMineR}}$.
\end{itemize}

We believe that the RGCCA package will be a valuable resource for researchers 
and practitioners who are interested in multiblock data analysis to gain 
new insights and improve decision-making.

The RGCCA framework is constantly evolving and extending. Indeed, we
proposed RGCCA for multigroup data \citep{Tenenhaus2014b}, RGCCA for
multiway data \citep{Gloaguen2020, Girka2023} and RGCCA for (sparse and
irregular) functional data \citep{Sort2023}. In addition, maximizing
successive criteria may be seen as sub-optimal from an optimization point
of view, where a single global criterion might be preferred. A global
version of RGCCA \citep{Gloaguen2020b}, which allows simultaneously extracting
several components per block (no deflation procedure
required), has been proposed. Also, it is possible to use RGCCA in
structural equation modeling with latent and emergent variables for
obtaining consistent and asymptotically normal estimators of the
parameters \citep{Tenenhaus2023}. At last, several alternatives for
handling missing values are discussed in \cite{Peltier2022}. Work in
progress includes the integration of all these novel approaches in the
next release of the RGCCA package.

# Acknowledgments {-}

This project has received funding from UDOPIA - ANR-20-THIA-0013, the European 
Union’s Horizon 2020 research and innovation program under grant 
agreement No 874583 and the AP-HP Foundation, within the framework of the 
AIRACLES Chair.

# References
